---
title: "Trait prioritization and crop performance with decentralized on-farm trial data"
package: gosset
author:
- name: Kauê de Sousa
  affiliation: Department of Agricultural Sciences, Inland Norway University, Hamar, Norway </br> Systems Tranformation, Bioversity International, Montpellier, France 
output: html_document
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  % \VignetteDepends{nasapower}
bibliography: ["vignette.bib"]
csl: citation_style.csl
---

# Summary

The gosset package provides a set of tools and methods to implement a workflow to analyse experimental agriculture data, from data synthesis to model selection and visualisation. The package is named after W.S. Gosset aka ‘Student’, a pioneer of modern statistics in small sample experimental design and analysis.

In this example I show one of the possible workflows to assess trait prioritization and crop performance using decentralized on-farm data generated with the tricot approach[@vanEttentricotguide]. I use the `nicabean` data from gosset. This data set was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot approach, farmers were asked to test three varieties of common bean randomly assigned as incomplete blocks of three varieties (out of 10 varieties) and assess which of those three had the best and worst performance in nine traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste, and overall appreciation).

# Trait prioritization 

First I load the packages and the data. The `nicabean` data is a list with two data frames, the first `nicabean[["bean_covar"]]` contains the covariates associated with the data, the second `nicabean[["bean_rank"]]` contains the trial data with farmers' evaluations. For better handling we take both data frames out of the list as separated data frames.

```{r fetch, message=FALSE, eval=FALSE, echo=TRUE}
library("PlackettLuce")
library("gosset")
library("climatrends")
library("nasapower")

data("nicabean", package = "gosset")

dat <- nicabean$bean_rank

covar <- nicabean$bean_covar

```

Farmers assessed nine traits, with overall appreciation, being a trait that describes farmers' overall evaluation considering all the performance of varieties over the season. The trait is a mix between farmers' appreciation (choice behaviour) and the "real" on-farm variety performance. Using the evaluations of other traits, which are more oriented to the trait performance, we can assess the drivers of farmers' appreciation. For this, I use the function `kendallTau()`, which measures the rank correlation between two traits, with overall appreciation as baseline. 

First, I transform the rankings into Plackett-Luce rankings[@Turner2020] using the function `rank_numeric()`. I run iterativelly over the traits adding the rankings to a list called `R`.

```{r , message=FALSE, eval=FALSE, echo=TRUE}

```



```{r , message=FALSE, eval=FALSE, echo=TRUE}
# build the rankings and put into a list
pldat <- list()

# run over the rankings of each trait
for(i in seq_along(traits)){
  
  # select the item names and rankings for the trait in the iteration i
  d_i <- dat[, c(items, paste0(traits[i], c("_best", "_worst")))]
  # not observed as NA
  d_i[d_i == "Not observed"] <- NA
  # check for ties in the response pos == neg
  keep <- d_i[,4] != d_i[,5] & !is.na(d_i[,4]) & !is.na(d_i[,5])
  # keep only the TRUE values out of the later validation
  d_i <- d_i[keep, ]
  
  names(d_i)[4:5] <- c("best", "worst")
  
  R_i <- rank_tricot(d_i, items = 1:3, input = c("best", "worst"))
  
  pldat[[i]] <- R_i
  
}

pldat

# fit the PlackettLuce model 
mod <- lapply(pldat, PlackettLuce)

```

The function `worth_map()` is a visualization tool to help in identifying variety performance based on different traits. It plots the log-worth (worth in log scale) of each trait indicating the variety performance. Positive values indicate a superior performance for a given trait, while negative values a under performance.

```{r , message=FALSE, eval=FALSE, echo=TRUE}
worth_map(mod, labels = traits)
```


# Crop performance with environmental data

Here I use the overall performance of varieties combined with environmental data to to assess crop variety performance. This approach was also used by van Etten et al. (2019)[@vanEtten2019]. I use the function `temperature()` from the climatrends[@climatrends] package to compute temperature indices for the first 80 days after planting in each data point. The climate data used is NASA POWER data supported by the package nasapower[@Sparks2018].

```{r , message=FALSE, eval=FALSE, echo=TRUE}
temp <- temperature(dat[, c("lon","lat")], 
                    day.one = dat[, "planting_date"],
                    span = 80)
```

Then I build the rankings using the function `rank_tricot()` but now as a `grouped_rankings` by adding the argument group = TRUE. This enables the ranking to be linked to covariates, in that case the temperature indices. Then with the function `pltree()` I fit a Plackett-Luce tree using the maximum night temperature (maxNT) and maximum day temperature (maxDT).

```{r , message=FALSE, eval=FALSE, echo=TRUE}
R <- rank_tricot(dat, 
                 items = c("variety_a","variety_b","variety_c"), 
                 input = c("overall_best","overall_worst"),
                 group = TRUE)

pld <- cbind(R, temp)

pl <- pltree(R ~ maxNT + maxDT, 
             alpha = 0.1,
             gamma = TRUE,
             data = pld)
```

The sets of functions below help in the visualization and identification of varieties with better performance in each node of the tree. 

```{r , message=FALSE, eval=FALSE, echo=TRUE}
plot(pl)

node_rules(pl)

top_items(pl, top = 5)

worst_regret(pl)

worth_map(pl)
```

## References

